---
title: "summarize practice, `count()`, `across()`"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    toc: yes
    toc_float: yes
    keep_md: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Produce clear, concise summaries using a variety of functions in `dplyr` and `janitor.`  
2. Use the `across()` operator to produce summaries across multiple variables.  

## Breakout Rooms  
Please take 5-8 minutes to check over your answers to the HW in your group. If you are stuck, please remember that you can check the key in [Joel's repository](https://github.com/jmledford3115/BIS15LW2021_jledford).  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("skimr")
library("palmerpenguins")
```

## Review
The summarize() and group_by() functions are powerful tools that we can use to produce clean summaries of data. Especially when used together, we can quickly group variables of interest and save time. Let's do some practice with the [palmerpenguins(https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data to refresh our memory.
```{r}
glimpse(penguins) #look at info about the df
```
```{r}
levels(penguins$island) #what are the islands? regardless of count what will we find in here?
```
```{r}
levels(penguins$species) #same but with species, we are looking at a lot of data for only a few species
```

As biologists, a good question that we may ask is how do the measured variables differ by island (on average)?
```{r}
penguins %>% 
  group_by(island) %>% #only want to look at info when grouped by name of island
  summarize(mean_body_mass_g=mean(body_mass_g), # mean of body mass on each island
            n=n())
```

Why do we have NA here? Do all of these penguins lack data?
here we are trying to see how many NAs there are in these data. keep track of MAs! then filter out NAs
```{r}
penguins %>% # I only want to look at the info in this dataframe when these groups of islands are squished together
  group_by(island) %>% 
  summarize(number_NAs=sum(is.na(body_mass_g))) #summarize how many(sum) NAs(is.na) are in the column body_mass_g for each island(group_by).
```

Well, that won't work so let's remove the NAs and recalculate.
here is where we filter out the NAs, nothing that is and NA in the column body_mass_g, then give me the mean of the data. 
```{r}
penguins %>% 
  filter(!is.na(body_mass_g)) %>% #removing NAs
  group_by(island) %>% #looking at only islands, we used level to get this
  summarize(mean_body_mass_g=mean(body_mass_g), # mean of each category when grouped by island but more than one category this time
            sd_body_mass_g=sd(body_mass_g),
            n=n())
```

What if we are interested in the number of observations (penguins) by species and island?
```{r}
penguins %>% 
  group_by(island, species) %>% #you can apparently group by more than 1 thing. torgerson only has adelie penguins on it, gentoo are only on Biscoe island
  summarize(n=n(), .groups= 'keep') #the .groups argument here just prevents a warning message #n=n() n is the name of a new column. this is counting how many penguins are in each group
```

## Counts
Although these summary functions are super helpful, oftentimes we are mostly interested in counts. The [janitor package](https://garthtarr.github.io/meatR/janitor.html) does a lot with counts, but there are also functions that are part of dplyr that are useful.  

`count()` is an easy way of determining how many observations you have within a column. It acts like a combination of `group_by()` and `n()`.
```{r}
penguins %>% 
  count(island, sort = T) #sort=T sorts the column in descending order
```

Compare this with `summarize()` and `group_by()`.
```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(n=n())
```

You can also use `count()` across multiple variables.
```{r}
penguins %>% 
  count(island, species, sort = F) #easier way to count
```

For counts, I am starting to prefer `tabyl()`. Lots of options are supported in [tabyl](https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html)
```{r}
penguins %>% 
  tabyl(species, island) #another bomb ass way to count stuff
```

```{r}
penguins %>% # this looks like a wa yto make the data more fancy and give a bit more info
  tabyl(species, island) %>% 
  adorn_percentages() %>% 
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns()
```

## Practice
1. Produce a summary of the mean for bill_length_mm, bill_depth_mm, flipper_length_mm, and body_mass_g within Adelie penguins only. Be sure to provide the number of samples.
# question: how would you add body mass back in without NAs? look this up
```{r}
penguins %>%
  filter(species=="Adelie") %>% # from the dataset, penguins, just look at info for species(column) Adelie(value). filteR=rows
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%  # the only colums we are interested in are these here. seleCt=columns
  summarize(mean_bill_length_mm=mean(bill_length_mm, na.rm=T), 
            mean_bill_depth_mm=mean(bill_depth_mm, na.rm =T), 
            mean_flipper_length_mm=mean(flipper_length_mm, na.rm =T), 
            mean_body_mass_g=mean(body_mass_g, na.rm=T)) # summarize lets us use different statistics to get info out. here we are wanting the mean of some columns and wew are making whole new columns based on the answers
```
2. How does the mean of `bill_length_mm` compare between penguin species?
```{r}
penguins %>% 
  group_by(species) %>% 
  summarize(n=n(), .groups= 'keep')
```

3. For some penguins, their sex is listed as NA. Where do these penguins occur?
```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(number_NAs=sum(is.na(sex)))
```

## `across()`
Last time we had some great questions on how to use `filter()` and `select()` across multiple variables. There is a new function in dplyr called `across()` which is designed to work across multiple variables. Have a look at Rebecca Barter's awesome blog [here](http://www.rebeccabarter.com/blog/2020-07-09-across/) as I am following her below.  

What if we wanted to apply summarize in order to produce distinct counts over multiple variables; i.e. species, island, and sex? Although this isn't a lot of coding you can image that with a lot of variables it would be cumbersome.
```{r}
penguins %>%
  summarize(distinct_species = n_distinct(species),
            distinct_island = n_distinct(island),
            distinct_sex = n_distinct(sex))
```

By using `across()` we can reduce the clutter and make things cleaner. 
```{r}
penguins %>%
  summarize(across(c(species, island, sex), n_distinct))
```

This is very helpful for continuous variables.
```{r}
penguins %>%
  summarize(across(contains("mm"), mean, na.rm=T))
```

`group_by` also works.
```{r}
penguins %>%
  group_by(sex) %>% 
  summarize(across(contains("mm"), mean, na.rm=T))
```

Here we summarize across all variables.
```{r}
penguins %>%
  summarise_all(n_distinct)
```

Operators can also work, here I am summarizing `n_distinct()` across all variables except `species`, `island`, and `sex`.
```{r}
penguins %>%
  summarise(across(!c(species, island, sex), 
                   n_distinct))
```

All variables that include "bill"...all of the other dplyr operators also work.
```{r}
penguins %>%
  summarise(across(starts_with("bill"), n_distinct))
```

## Practice
1. Produce separate summaries of the mean and standard deviation for bill_length_mm, bill_depth_mm, and flipper_length_mm for each penguin species. Be sure to provide the number of samples.  
```{r}
penguins %>%
  group_by(species) %>% 
  summarize(across(contains("mm"), mean, na.rm=T), n=n())
penguins %>%
  group_by(species) %>% 
  summarize(across(contains("mm"), sd, na.rm=T), n=n())
```


## Wrap-up  

Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)